|

### 2025-05-28 Reviews/round_1/rebuttal_point_by_point.md

**In the Methods section and in the Supplementary Materials, the experimental procedure is largely described in a comprehensive and reproducible manner. However, the methods/experiments/questionnaire items are not related to the previously stated research questions or hypotheses. This makes it unnecessarily difficult for the reader to assess the validity of the approach chosen to address the research questions. To enhance comprehensibility, it would be beneficial to explicitly label the independent and dependent variables, as well as the control variables collected for each research question/hypothesis.**

> We fully agree with this point. In the Results section, we now make a direct reference linking the analysis to each of the five research questions.

**The results section first describes the group of subjects. For one experimental intervention (socratic AI tutor), it is reported whether the experimental group differed significantly from the control group. However, this information is missing for the other experimental intervention (AI explanation). In the remainder of the section, the analyses of the dependent variables are presented in a largely scientific manner. Notably, for the first experimental manipulation, a dependent variable has been listed in the methods section (propensity to update initial guess). However, no results are reported for this variable, and I do not understand what was done with this data or why it wasn't used. In the second experimental manipulation (socratic tutor), variables are reported that have not been described in the methods section. These new variables that measure the interaction with the AI tutor are only (and very briefly) motivated here. They should be motivated in the Introduction and included in the Methods section.**

> We thank the reviewer for these suggestions that greatly improved the clarity of our methods. Specifically: 
> 
> - Regarding the descriptive information about the other experimental intervention, Table 2 now shows summary statistics for both treatment groups. It also shows p-values from separate statistical tests of independence between variable and treatmnet assignment.
> 
> - Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity. Accordingly, we have included the following sentence in the Results section (p. 16): *Regarding the propensity to update the guess, we found no evidence that the treatment had an effect on this propensity: Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher’s test, p = 0.9). These revisions showed no systematic pattern, as there was no evidence that the AI explanation influenced the direction or magnitude of the changes.*

> - About the student-AI interaction metrics for the Socratic intervention that have not been fully described in the Methods section, we now report this information in a dedicated sub-section titled "Student-AI Interaction Metrics" in the Methods section (p. 14). We report the main paragraph below: 
> *To quantify engagement, we analyzed students’ interaction logs with the AI Tutor, recording the number of turns and the word counts as a proxy for interaction intensity. While this provides a basic behavioral metric, it does not capture the quality of cognitive engagement or helpfulness,
limiting its direct relevance to RQ1 and RQ2. Therefore, we also asked students about how useful they found the AI interactions and how confident they were in their answers to selected tasks. These self-reported metrics are needed to assess whether explanations were helpful to students (for RQ1) and whether Socratic dialogue promoted critical thinking or confidence (for RQ2).*

### 2025-05-28 01-intro.Rmd
edits + rebuttals

A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries [@wei2022chain]. Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions;@danovitch2021mind], there is limited empirical research examining how such AI-generated explanations impact students' learning processes. Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills. To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) to address our first research question (RQ):

**RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks?**

>>>>

A key advancement over previous AI systems is that LLMs can provide step-by-step reasoning to answer students' queries [@wei2022chain]. Although evidence suggests that such explanations can foster children's learning and contribute to the development of their scientific reasoning [@dejong1986explanation;@legare2014contributions;@danovitch2021mind], there is limited empirical research examining how such AI-generated explanations impact students' learning processes. Most prior studies have not isolated the specific role of detailed AI-generated reasoning in enhancing problem-solving skills. To fill this gap, we conducted a randomised experiment with k-12 students (n = 122) who were engaged in a simple estimation task -- guessing the value of coins in a jar. Half of the participants received only an AI-generated prediction, while the other half were additionally provided with AI-generated explanations detailing how to derive such estimation. This setup enabled to address our first research question (RQ):

**RQ1: Do AI-generated explanations enhance students' problem-solving performance in school tasks, specifically in terms of estimation accuracy?**

### 2025-05-28 03-results.Rmd
edits + rebuttals

@@

We tested whether showing students AI-generated step-by-step reasoning affected how they used the AI's prediction (RQ1), compared to just seeing the prediction without any reasoning. For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s guess, $\text{Guess}_i$, and the actual value of coins:
$$
  \text{Absolute Error}_i = \mid \text{Guess}_i - \text{Actual Value Coins} \mid,
$$
where a smaller absolute error indicated greater accuracy.

>>>>

```{r update-propensity, include = FALSE}
library(boot)
library(MASS)

names(tutorgpt)

# data 
update_ds <- tutorgpt %>% 
  mutate(update = sign(q12_guess_coins_final - q9_guess_coins_initial),
         initial_guess_100 = q9_guess_coins_initial / 100)

update_propensity_table <- update_ds %>% 
  xtabs(formula = ~ update  + treatment_explain_id)

knitr::kable(update_propensity_table)
fisher.test(update_propensity_table)

# util functions
center <- function(x) scale(x, center = TRUE, scale = FALSE)
topcode <- function(x, value) ifelse(x > value, value, x)

# Original fit
model <- center(q12_guess_coins_final) ~ initial_guess_100 +
  I(initial_guess_100^2) + treatment_explain_id + location

# Fit regular ols
fit_update_propensity <- lm(model, update_ds)

# Robust regression
fit_robust <- rlm(model, update_ds)

# remove outliers
fit_no_out <- fit_robust %>% 
  update(subset = q9_guess_coins_initial < 1e5)

fit <- list(fit_update_propensity, fit_robust, fit_no_out)
stargazer::stargazer(fit, type = "text", keep.stat = "n", digits = 1)

# ---- Figure ----
cap <- "Difference between first and second guess.  "
tutorgpt %>% 
  mutate(
    final_tc = q12_guess_coins_final %>% 
      topcode(1e3),
    initial_tc = q9_guess_coins_initial %>% 
      topcode(1e3),
  ) %>%
  ggplot(aes(x = initial_tc, 
             y = final_tc,
             color = treatment_explain_id,
             pch = treatment_explain_id,
             fill = treatment_explain_id,
             linetype = treatment_explain_id)) +
  geom_point() +
  geom_smooth(alpha = .25) + 
  #scale_y_log10() + scale_x_log10() +
  labs(
    x = "First Guess ($)",
    y = "Second Guess ($)"
  ) + 
  theme(
    legend.position = "bottom",
  )

@@

Nearly all students revised their initial estimates (110 out of 122), with no significant difference between treatment groups (Fisher’s test, p = 0.9). These revisions showed no systematic pattern, as there was no evidence that the AI explanation influenced the direction or magnitude of the changes. However, presenting students with AI-generated step-by-step reasoning may have influenced the accuracy of their revised estimates (RQ1).

To address RQ1, we examined whether the treatment—showing step-by-step AI reasoning—affected how students used the AI’s prediction, compared to seeing the prediction alone without any accompanying explanation. For each student $i = 1, \cdots, 122$, we computed the absolute error as the absolute difference between student $i$'s guess, $\text{Guess}_i$, and the actual value of coins:
$$
  \text{Absolute Error}_i = \mid \text{Guess}_i - \text{Actual Value Coins} \mid,
$$
where a smaller absolute error indicated greater accuracy.
